A Computational Account of Default-Mode Function\\ by Control Theory and Reinforcement Learning

### 2016-09-29 paper/paper.tex
a pass on RL section

\section{Reinforcement-Learning as a computational model for DMN function}
\subsection{General Reinforcment-Learning (RL) theory}
In a general reinforcement-learning (RL) framework, an \textit{agent} interacts with an unknown
environment by taking \textit{actions / decisions} $a_t$ which trigger changes in the
\textit{state} of the environment
$s_t \rightarrow s_{t+1}$, accompanied with a real-valued \textit{reward} $r_t$ (or regret!,
since it can be
negative). This should be thought of as an instantaneous satisfaction accompanying the execution of
an action. Here the subscript $t$ denotes time, and can be discrete or continuous.
%% The unknown environment $\mathcal E$ is generally taken as stochastic, but having some kind of
%% memory.
A popular abstract model for the scenario are so-called \textit{Markov Decision Processes (MDPs)}.
This is simply a quintuplet $(\mathcal S, \mathcal A, r, \rho, \mu)$ where
\begin{itemize}
\item $\mathcal S$ is the set of states, e.g $\mathcal S = \{\text{happy}, \text{sad}\}$.
\item $\mathcal A$ is the set of actions, e.g $\mathcal A = \{\text{read}, \text{run},
  \text{laugh}, \text{empathize}, \text{empathize}\}.$
\item $r : \mathcal S \times \mathcal A \rightarrow \mathbb R$, so that $r(s, a)$ is the instant reward for taking action $a$ at state $s$.
\item $\mu: \mathcal S \rightarrow [0, 1]$ is the prior probability on the states so that
  $\mu(s)$ is the probability that the environment starts-off in state $s$.
  \item $\rho : \mathcal S \times \mathcal A \times \mathcal S \rightarrow [0, 1]$, so that
    $\rho(s,a,s') = \mathbb P(s'|s, a)$ is the probability of moving to state $s'$ if action $a$ is taken from state $s$. In addition, one requires that such transitions be Markovian.
\end{itemize}

@@

Since an action taken now might have repercussions long into the feature, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s_t, a_t)$, a cumulative version which
takes into account the future. A standard for modeling this accumulation is the so-called
time-discounted \textit{cumulative reward} function
\begin{equation}
  \label{eq:cumr}
  R(s_t, a_t) = r(s_t, s_t) + \gamma r(s_{t+1}, s_{t+1}) +\ldots + \gamma^Tr(s_{t+T}, s_{t+T})
  = r(s_t, s_t) + \gamma \sum_{\tau=1}^{T}\gamma^{\tau-1}r(s_{t+\tau},a_{t+\tau}),
  \end{equation}
for some \textit{discount factor} $\gamma \in [0, 1]$, and horizon $T$. The choice $\gamma = 0$ (or $T=0$)  corresponds to a perfectly myopic agent who is solely concerned about their immediate rewards and have no regards for the future. For there to be any learning at all, a technical requirement is that $\gamma < 1$ \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
Thus the agent is positively biased towards considering rewards which are will
come soon. Note that $R(s_t, a_t)$ defined above is a stochastic quantity which depends on the
trajectory of a policy $\pi$,
\begin{eqnarray*}
  \begin{split}
    s_0 &\sim \mu\\
    a_0 &\sim \pi(.|s_0)\\
    s_1 &\sim \rho(.|s_0,a_0) \text{ and collect rewards }r(s_0, a_0)\\
    a_1 &\sim \pi(.|s_1)\\
    s_2 &\sim \rho(.|s_1,a_1) \text{ and collect rewards }r(s_1, a_1)\\
    \ldots\\
    a_t &\sim \pi(.|s_t)\\
    s_{t+1} &\sim \rho(.|s_t,a_t) \text{ and collect rewards }r(s_t, a_t)\\
    \ldots
  \end{split}
\end{eqnarray*}

The goal is then to sample from a policy $a_t \sim \pi$ so that on average, $R(s_t, a_t)$ is maximized. Formally, the agent's goal is to maximize the $Q$-value at $(s_t, a_t)$, defined by
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s_t,a_t) = \mathbb E [R(s_t, a_t)]
\end{equation}

where the expectation is over all possible trajectories from $(s_t, a_t)$, under the policy $\pi$.

By construction, in general, maximizing the expectation in  \eqref{eq:qval} is intractable due to
the enormous number of possible trajectories to consider.
One popular simplification that removes this issue is to only consider the sub-family of \textit{deterministic policies}, which map each state to the best single best action which should be taken at that state, and therefore define \textbf{functions from states to actions}.

\textit{Q-learning} (Watkins \& Dayan, 1992), a commonly used off-policy algorithm, uses the
\textit{greedy policy}
\begin{equation}
  \pi(s) = \pi_{\tilde{Q}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a),
  \label{eq:qlearning}
\end{equation}
where $\tilde{Q}(s, a)$ is the current estimate of the state-action value table.
The hallmark of this trick wherein one only considers deterministic policies is the so-called
\textit{Bellman's equation}
\begin{equation}
  Q^\pi(s_t, a_t) = \mathbb E_{s_{t+1} \sim \rho(.|s_t)} [r(s_t,a_t) + \gamma Q^\pi(s_{t+1}, \pi(s_{t+1}))]
  \label{eq:bellman}
\end{equation}
which is a fixed-point equation for the deterministic policy $\pi$.
Notice that in the expectation in \eqref{eq:bellman}, the sampling is done over only things which
depend on the environment, and so can be learned off-policy by observing state-transitions
triggered by another behavioral policy, which can be stochastic.

@@

\paragraph*{Some notes:}
\begin{itemize}
\item The brain does backprop (see Hinton's recent talk at Stanford)
\item model weights $\theta^Q$ should correspond to connections between neurons
\item humm, looks like such a knowledge would simply translate into constraints on the DqN weights $\theta^Q$ (... at least, part of such knowledge)
  ok, gotta think about the model-free / model-based reconsiliation
\end{itemize}

>>>>

\section{Reinforcement-Learning as a computational model for DMN function}
\subsection{General Reinforcment-Learning (RL) theory}
In a general reinforcement-learning (RL) framework, an \textit{agent}
interacts with an unknown environment in trial-and-error fashion by taking
\textit{actions / decisions} $a_t$ which trigger changes in the
\textit{state} of the environment
$s_t \rightarrow s_{t+1}$, accompanied with a real-valued \textit{reward}
$r_t$ (or regret!, since it can be negative). This reward should be thought
of as an instantaneous satisfaction accompanying the execution of
an action. Here the subscript $t$ denotes time, and can be discrete or
continuous.
%% The unknown environment $\mathcal E$ is generally taken as stochastic, but having some kind of
%% memory.
A popular abstract model for the scenario are so-called \textit{Markov Decision Processes (MDPs)}.
This is simply a quintuplet $(\mathcal S, \mathcal A, r, \rho, \mu)$ where
\begin{itemize}
\item $\mathcal S$ is the set of states, e.g $\mathcal S = \{\text{happy}, \text{sad}\}$.
\item $\mathcal A$ is the set of actions, e.g $\mathcal A = \{\text{read}, \text{run},
  \text{laugh}, \text{empathize}, \text{empathize}\}.$
\item $r : \mathcal S \times \mathcal A \rightarrow \mathbb R$, so that $r(s, a)$ is the instant reward for taking action $a$ at state $s$.
\item $\mu: \mathcal S \rightarrow [0, 1]$ is the prior probability on the states so that
  $\mu(s)$ is the probability that the environment starts-off in state $s$.
  \item $\rho : \mathcal S \times \mathcal A \times \mathcal S \rightarrow [0, 1]$, so that
    $\rho(s,a,s') = \mathbb P(s'|s, a)$ is the probability of moving to state $s'$ if action $a$ is taken from state $s$. In addition, one requires that such transitions be Markovian.
\end{itemize}

@@

Since an action taken now might have repercussions long into the feature, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s_t, a_t)$, a cumulative version which
takes into account the future. A standard for modeling this accumulation is the so-called
time-discounted \textit{cumulative reward} function
\begin{equation}
  \label{eq:cumr}
  R(s_t, a_t) = r(s_t, s_t) + \ldots + \gamma^Tr(s_{t+T}, s_{t+T})
  = r(s_t, s_t) + \gamma \sum_{\tau=1}^{T}\gamma^{\tau-1}r(s_{t+\tau},a_{t+\tau}),
  \end{equation}
for some \textit{discount factor} $\gamma \in [0, 1]$, and horizon $T$. The choice $\gamma = 0$ (or $T=0$)  corresponds to a perfectly myopic agent who is solely concerned about their immediate rewards and have no regards for the future. For there to be any learning at all, a technical requirement is that $\gamma < 1$ \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
Thus the agent is positively biased towards considering rewards which are will
come soon. Note that $R(s_t, a_t)$ defined above is a stochastic quantity which depends on the trajectory generated by the policy $\pi$, upto but not including time $t$
\begin{eqnarray*}
  \begin{split}
    s_0 &\sim \mu\\
    a_0 &\sim \pi(.|s_0)\\
    s_1 &\sim \rho(.|s_0,a_0) \text{ and collect rewards }r(s_0, a_0)\\
    a_1 &\sim \pi(.|s_1)\\
    s_2 &\sim \rho(.|s_1,a_1) \text{ and collect rewards }r(s_1, a_1)\\
    \ldots\\
    a_{t-1} &\sim \pi(.|s_{t-1})\\
    s_{t} &\sim \rho(.|s_{t-1},a_{t-1}) \text{ and collect rewards }r(s_{t-1}, a_{t-1})
  \end{split}
\end{eqnarray*}

%% The goal is then to sample from a policy $a_t \sim \pi$ so that the expected
%% value of $R(s_t, a_t)$ is maximized.
Formally, the agent's goal is then to maximize the state-action value for
taking action $a_t$ and state $s_t$ defined by
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s_t,a_t) = \mathbb E [R(s_t, a_t)]
\end{equation}

where the expectation is over all possible trajectories successor
trajectories of state $s_t$.

By construction, in general, maximizing the expectation in  \eqref{eq:qval} is intractable due to
the enormous number of possible trajectories to consider.
One popular simplification that removes this issue is to only consider the sub-family of \textit{deterministic policies}, which map each state to the best single best action which should be taken at that state, and therefore define functions from states to actions.

\textit{Q-learning} (Watkins \& Dayan, 1992), a commonly used off-policy algorithm, uses the
\textit{greedy policy}
\begin{equation}
  \pi(s) = \pi_{\tilde{Q}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a),
  \label{eq:qlearning}
\end{equation}
where $\tilde{Q}(s, a)$ is the current estimate of the state-action value table.
The hallmark of this trick wherein one only considers deterministic policies is the so-called
\textit{Bellman's equation} [Sutton and Barto 1998]
\begin{equation}
  Q^\pi(s_t, a_t) = \mathbb E_{s_{t+1} \sim \rho(.|s_t)} [r(s_t,a_t) + \gamma Q^\pi(s_{t+1}, \pi(s_{t+1}))]
  \label{eq:bellman}
\end{equation}
which is a fixed-point equation for the deterministic policy $\pi$.
Notice that in the expectation in \eqref{eq:bellman}, the sampling is done over only things which
depend on the environment, and so can be learned off-policy by observing state-transitions
triggered by another behavioral policy, which can be stochastic.

@@

\paragraph*{Some notes:}
\begin{itemize}
  \item Actions are continuous
  \item The brain does backprop (see Hinton's recent talk at Stanford)
  \item model weights $\theta^Q$ should correspond to connections between neurons
\item humm, looks like such a knowledge would simply translate into constraints on the DqN weights $\theta^Q$ (... at least, part of such knowledge)
  ok, gotta think about the model-free / model-based reconsiliation
\end{itemize}
